Fixing Vulnerabilities Hinders Maintainability

### 2019-02-03 paper/ICPC19/icpc19.tex
comments macros
lista boring + refs and cites

\newcommand{\Sof}[1]{\textbf{[Sofia:[}{\color{red} #1}\textbf{]]}}

@@

\begin{abstract}
Security is a requirement of utmost importance to produce software with high
quality. However, there is still a considerable amount of zero-day
vulnerabilities being discovered, and fixed, almost weekly. We suspect that when
developers address these vulnerabilities on their codebases, they actually
affect the maintainability of software, potentially introducing other
vulnerabilities. This paper evaluates the impact of refactorings to improve
security on the maintainability of open-source software. Maintainability is
measured based on the \emph{Better Code Hub}'s model of 10 guidelines using a
dataset including $607$ security commits. Results show that fixing software
vulnerabilities hinders the maintainability of open-source software. This is
particularly the case for refactorings that address \emph{Broken Authentication
and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
attacks. Furthermore, we conclude that changes to codebases to fix vulnerabilities
need to be performed with extra care.
\end{abstract}

@@

This paper is structured as follows: section \ref{sec:motivation} introduces an
example of a security refactoring of a known vulnerability found in
OpenSSL\footnote{Repository available at https://github.com/openssl/openssl
(Accessed on \today{})}; section \ref{sec:methodology} describes the
methodology used to answer the research questions; section \ref{sec:results}
presents the results and section \ref{sec:discussion} discusses their
implications; section \ref{sec:threats}, threats to the validity of the study
are enumerated; section \ref{sec:rw} describres the different work and existing
literature in the field of study; and, finally section \ref{sec:conclusions}
concludes the main findings and elaborates on future work.
%
\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security expertise, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. It turns out that fixing these issues is as simple as
modifying the software codebase. However, these refactorings may have a negative
impact on software maintenance, mostly because developers follow the quickest
solution to fix the problem and not necessarily the most elegant and performant
one. Thus, we want to understand how security refactorings impact the
maintainability of software.

@@

\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
inserted to limit the number of bytes allowed in a DTLS handshake message for
\texttt{s}. This function replaces lines $18$---$21$.

\ref{lst:func2} Checks if the maximum certificate list size
(\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
(\texttt{max\_len}). If that condition holds, the function returns the size of
the certificate list, otherwise it returns the maximum length.

\ref{lst:func5} and \ref{lst:func4} add the check to ensure that the message
length is never higher than the maxiumum allowed by the DTSL protocol in
different functions.

Although limiting the size of a handshake message seems an elementary problem to
repair, the example shows that there was a significant amount of changes
performed in the codebase which produced a negative impact in the project
maintainability. Even though the developer tries to simplify the codebase with
the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), it
also disrupts $2$ of the guidelines proposed by the Software Improvement Group
(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}. It adds a
new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which contributes
to a potential violation of the
\emph{Keep Your Codebase Small} guideline. Hence, we have observed that the
impact of this refactoring on the OpenSSL maintainability was negative.

@@

There are security flaws that are more difficult to refactor than others. For
instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability with an \texttt{escape}
function. Knowing which patterns are more likely to increase maintainability
issues is one step forward to bring awareness to security engineers of what
patterns need more attention. The taxonomy of security patterns used to answer
this question is the same as the one provided by the authors of the dataset~\cite{Reis:2017:IJSSE}.
Maintainability is measured separately for each pattern.
%
\section{Methodology}\label{sec:methodology}
%
The methodology used to measure the impact of security refactorings on the
maintainability of open-source software is discussed this section. The methodology
comprehends the following steps, as illustrated in Figure \ref{fig:met}:
\begin{enumerate}
	\item Extract relevant data from a dataset containing
$716$ security refactorings collected from open-source software available on
GitHub~\cite{Reis:2017:IJSSE}.
	\item A baseline of regular commits was
randomly collected from the list of projects of the main dataset to evaluate the impact of regular commits on the maintainability of open-source
software.
\item Use the Software Improvement Group's\footnote{SIG's website is available
at https://www.sig.eu/ (Accessed on \today{})} web-based source code analysis
service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
https://bettercodehub.com/ (Accessed on \today{})} to quantify
maintainability for both security and regular commits.
\end{enumerate}

%
\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
 	\caption{Study Methodology}
	\label{fig:met}
\end{figure}
%
\subsection{Dataset}
%
To evaluate the impact of security refactorings on the maintainability of
open-source software, we use a dataset of security flaws which is the outcome of
mining $94$ GitHub projects. Reis and Abreu (2017) mined open-source
software aiming at the extraction of real examples --- created by real
developers --- to test and assess the performance of static analysis tools since
using hand-seeded test cases or mutations could lead to misleading assessments
of the capabilities of the tools \cite{just2014mutants}. The study yielded a
dataset with $716$ test cases for $16$ security patterns. Each test case of the
dataset is a triplet: the commit before the refactoring, the commit responsible
for the refactoring, and the snippets of code that differ from one version to
another (typically, called \textit{diffs}) --- where one can easily review the
code used to fix the security flaw. In this study, we focus on computing the
maintainability of the commits before and after the security refactoring to
evaluate if the impact was positive, negative or none.

@@

The test cases in the dataset are classified with one of the $16$ following
patterns: \textit{Injection}, \textit{Broken Authentication
and Session Management}, \textit{Cross-Site Scripting}, \textit{Broken Access
Control}, \textit{Security Misconfiguration}, \textit{Sensitive Data Exposure},
\textit{Insufficient Attack Protection}, \textit{Cross-Site Request Forgery},
\textit{Using Components With Known Vulnerabilities}, \textit{Unprotected APIs},
\textit{Memory Leak}, \textit{Overflow}, \textit{Denial-of-Service},
\textit{Path Traversal}, \textit{Miscellaneous}, \textit{Resource Leak} and
\textit{SHA-1 Hash Function}. Some of the patterns are based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top 10 of 2017~\cite{oswap:2017}

@@

	\item \textbf{Denial-of-Service} Security flaws that can allow a cracker to
	flood or crashing services. This attacks usually occur when the system
	receives much traffic causing it to slow down and eventually stop. For
	example, flaws allowing crackers to trigger memory allocations corresponding
	to large length values (Listing \ref{lst:vuln}).

	\item \textbf{Miscellaneous} This pattern comprises several others categories
	of security refactorings (e.g., \textit{path traversal and buffer overflow}).
	It contains several other security refactorings that do not have their own
	pattern yet. The patterns not satisfying the size requirement of more than 20
	refactorings are contemplated here.

@@

The dataset contemplates security flaws for more than 13 different languages
being PHP (39\%) and C (24\%) the most prevalent ones; see Figure \ref{fig:lang}.
To classify the applications in the dataset, we use a taxonomy for open-source
software from previous work~\cite{7816479}. Figure \ref{fig:domain}
presents the projects domain distribution: \textit{Aplication Software} ($25$),
software that provides end-users with functional systems; \textit{Web libraries
and frameworks} ($20$); \textit{System Software} ($18$), software that provides
services and infrustructures (e.g., operating systems, servers and databases);
\textit{Software Tool} ($13$), software that supports development (e.g.,
programming languages, compilers, package managers, IDEs); and, \textit{Non-web
libraries and frameworks} ($18$), for desktop and mobile software.
%
Table \ref{tab:dataset} presents the descriptive statistics of the $94$
open-source projects involved in this study, including the number of forks, the number of
stars, the number of watchers, the number of contributors, the number of commits, the number of
branches, the number of releases,  size (cf. GitHub it is the size of the whole
repository, including all of its history, in kilobytes), the number of issues, and the
number of pull requests.

@@

As said before, the web-based source code analysis service \emph{Better Code
Hub} (BCH) is used to collect the maintanability reports of the refactorings of
each project. Table \ref{tab:guidelines} presents the 10 guidelines
proposed by BCH's authors for delivering software that is easy to maintain
\cite{Visser:2016:OREILLY}. All guidelines are computed in the same evaluation by
BCH. During each guideline evaluation, the tool determines the compliance towards
one guideline by establishing limits for the percentage of code allowed to be in
each of the $4$ risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk}
and \emph{very high risk}). If the project does not violate the thresholds, then it
is compliant with the guideline. These thresholds are calibrated by BCH using
their own data/experience --- open-source and closed software systems. If a
project is compliant with a guideline, it means that it is at least $65\%$
better than the software used by BCH to calculate the thresholds\footnote{Check
the answer to \emph{How can I adjust the threshold for passing/not passing a
guideline?} at https://bettercodehub.com/docs/faq (Accessed on \today{})}.

Figure \ref{fig:bchrep} shows an example of the report
provided by BCH for a project after finishing its evaluation. The examples
refers to the OpenSSL CVE-2014-3506 vulnerability refactoring ---
as described as motivating example in Section \ref{sec:motivation}. This
version of OpenSSL only complies with 1 out of 10 guidelines: \emph{Write
Clean Code}.

@@

\emph{Units} are the smallest groups of code that can be maintained and executed
independently \cite{Visser:2016:OREILLY}. McCabe \cite{1702388} is a software
metric that calculates the number of linearly independent paths of the source
code. BCH uses this metric to measure the number of branch points in the
\emph{Write Simple Units of Code} guideline. The bar represents the top 30 of
units that violates the guideline, sorted by severity, which is indicated by the
colors of the checkboxes. The green bar represents the number of compliant
branch points per unit (\emph{at most 5}), i.e., the units are compliant with
ISO 25010 \cite{iso:2011}. Yellow, orange and red bars represent units that do
not comply with medium (\emph{above 5}), high (\emph{above 10}) and very high
(\emph{above 25}) severity levels. In the bar, there are marks that pinpoint the
compliance thresholds for each severity level. If the green mark is somewhere
in the green bar it means it is compliant with a low level of severity.

@@

Aiming to analyze the impact of security refactorings, we use BCH to compute
maintainability of two different versions of the project (Figure
\ref{fig:commit}):
\begin{itemize}
	\item $v_{s-1}$, the version containing the security flaw, i.e., before the
	refactoring;
	\item $v_{s}$, the version after fixing the security flaw, i.e., after the
	refactoring;
\end{itemize}

The BCH tool does not compute the final score that our study needs to compare
maintainability amongst different project versions. Thus, our work proposes an
equation to capture the distance between the current state of the project and
the standard thresholds based on previous work \cite{Olivari:2018}.
\begin{itemize}
	\item \textbf{Size project changes do not affect the maintainability
	difference between two versions, $\Delta M (v_{s-1},v_{s})$.} In this work, we
	aim to evaluate the alike security patterns occurring in different projects.
	Thus, normalization based on its size cannot be used - we convert relative
	data to the respective number of lines of code.
	\item \textbf{Distance to lower severity levels are less penalized than to the
	thresholds in high severity levels.} Severity level weights based on the
	severity level to count lines of code that violate maintainability guidelines.
\end{itemize}

@@

\noindent
where $G$ is the group of selected maintainability guidelines from BCH (Table
\ref{tab:guidelines}) and $v$ is the version of the software under evaluation.
The maintenance $M$ based on the guideline $g$ for a given version of a project
is computed with the following equation:

@@

\noindent
where $T(l)$ is the threshold in percentage of the lines of code that are
accepted to be non-compliant with the guideline for the severity level $l$. This
is a standard value defined by BCH. In other words, the factor $w$ is used in
Equation \ref{eq:3} to highlight the lines of code that are not complying with
the guideline. Then, we compute the difference of maintainability between the
security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in
Figure \ref{fig:commit}.

\subsection{Statistical Validation}\label{sec:statsval}
%
To validate the maintainability differences in different groups of commits
(e.g., baseline and security commits), we use the Paired Wilcoxon signed-rank
test with the significance level $\alpha = 0.05$ \cite{10.2307/3001968} . In
other words, we test the null hypothesis that the maintainability difference
between pairs of versions $v_{v-1}$, $v_v$ (i.e., before and after a
security-commit) follows a symmetric distribution around $0$. However, when the
difference between groups is zero, the observations are discarded. Thus, we use
a more robust approach of the wilcoxon test, the Wilcoxon test by Pratt which
provides an alternative that incorporates the cases where the maintainability
difference is zero \cite{10.2307/2282543}. To understand the effect-size, as
advocated by the Common-language effect sizes \cite{graw:1992}, we compute the
mean difference, the median of the difference, and the percentage of cases that
reduce maintainability.
%
\section{Results}\label{sec:results}

In total, this study evaluates $607$ security and baseline commits from 94
different open-source projects distributed among 5 different main types of
software, Figure \ref{fig:domain}. The following section presents the results
obtained for each research question.

@@

The impact of security and baseline commits in the maintainability of
open-source software is presented in Figure \ref{fig:secvsreg}. For each type
of change, three types of impact of refactorings in the software maintainability
are presented: \emph{negative}, maintainability decreases (\emph{red}); \emph{none},
maintainability stays the same (\emph{yellow}); and, \emph{positive}, maintainability
increases (\emph{green}). In front of each group, it is presented the mean ($\overline{x}$)
and median ({$\widetilde{x}$) of the maintainability difference and the p-value
resulting from the Wilcoxon signed-rank test.

@@

In this study, we also evaluated the impact of each programming language on the
maintainability of open-source software, Figure \ref{fig:lang_main}. Although the
dataset contains security refactorings for more than $13$ languages, we only
present the results for \emph{Java}, \emph{C}, \emph{Ruby} and \emph{PHP} because
one of the requirements of the Wilcoxon signed-rank test is that the input sample
has to have more than $20$ measurements. The other languages were integrated in the
\emph{Others} set. The results of the Wilcoxon signed-rank test by pattern returned
statistical evidence which supports that the maintainability of open-source software
decreases for security refactorings in 3 different languages: \emph{C}
(p = $5.72$x$10^{-9}$), Ruby (p = 0.057) and PHP (p = $3.54$x$10^{-6}$). The results
also provide statistical significance for when security refactorings are performed
in \emph{Java} (p = $3.61$x$10^{-2}$). \emph{Python} and the other languages
that integrate the \emph{Others} set contain more cases where the maintainability
increases. However, the results obtained for these patterns are not statistically
significant.

@@

The impact of each type of security refactoring on the maintainability of
open-source software is presented in Figure \ref{fig:pat}. The results of the
Wilcoxon signed-rank test by pattern returned statistical evidence which
supports that the maintainability of open-source software decreases for 4
different patterns: Broken Autentication (p = 0.018), Denial-of-Service
(p = 0.029), Memory Leak (p = 0.009) and Miscellaneous (p = $3.07$x$10^{-4}$).
The results also returned statistical significance that evidence that maintainability
is more likely to not suffer differences for open-source software after removing
Components with Known Vulnerabilities (p = $7.96$x$10^{-4}$) and fixing Cross-Site
Scripting Vulnerabilities (p = 0.001). Injection and Cross-Site Request Forgery
produced more cases where the refactorings had a negative impact in the maintainability.
However, the results obtained for these patterns are not statistically significant.

@@

40.2\% of the security refactoring have a negative impact on the maintainability
of open-source software, Figure \ref{fig:secvsreg}. This raises a new tradeoff when
developers need to fix vulnerabilities on their projects because they might be affecting
the software maintainability or even introducing new vulnerabilities. This study exhibits
proof that developers may have to reduce maintainability for the sake of safety. We understand
that developers should be able to produce secure code without affecting the maintainability of
their projects. But there are still some concerns that need to be addressed:
\begin{itemize}
	\item \textbf{Security Frameworks} need to be secure and user-friendly~\cite{5287006, 7676144}. Unfortunately,
	there are several vulnerabilities that stay undisclosed. Thus, developers should not
	trust them blindly and check defaults.
	\item \textbf{Documentation} of software should provide developers with the best practices to implement safe and maintainable software. Unmaintainable code is difficult to test, analyze and re-use.
	\item \textbf{Proper Authorization Mechanisms} should be designed and implemented. This is
	one of the most complex security features to implement. Some mechanisms might assist authorization maintenance in long run, such as the combination of \texttt{HTTP} filters with \texttt{Spring Security} annotations.
	\item \textbf{Deprecated code} should be avoided by developers. They should integrate tools such as BCH on their projects CI/CD pipelines to help them be more aware of the maintainability of their code and issues (e.g., dead and duplicated code).
	\item\textbf{Programming Languages} should provide new mechanisms to easily implement security patterns without endangering maintainability.

>>>>

\newcommand{\Sof}[1]{\textbf{[Sofia:[}{\color{red} #1}\textbf{]]}}
\newcommand{\Luis}[1]{\textbf{[Luis:[}{\color{green} #1}\textbf{]]}}
\newcommand{\Rui}[1]{\textbf{[Rui:[}{\color{blue} #1}\textbf{]]}}

@@

\begin{abstract}
Security is a requirement of utmost importance to produce software with high
quality. However, there is still a considerable amount of zero-day
vulnerabilities being discovered, and fixed, almost weekly. We suspect that when
developers address these vulnerabilities on their codebases, they actually
affect the maintainability of software, potentially introducing other
vulnerabilities. This paper evaluates the impact of refactorings to improve
security on the maintainability of open-source software. Maintainability is
measured based on the \emph{Better Code Hub}'s model of 10 guidelines using a
dataset including 607 security commits. Results show that fixing software
vulnerabilities hinders the maintainability of open-source software. This is
particularly the case for refactorings that address \emph{Broken Authentication
and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
attacks. Furthermore, we conclude that changes to codebases to fix vulnerabilities
need to be performed with extra care.
\end{abstract}

@@

This paper is structured as follows: section~\ref{sec:motivation} introduces an
example of a security refactoring of a known vulnerability found in
OpenSSL\footnote{Repository available at https://github.com/openssl/openssl
(Accessed on \today{})}; section~\ref{sec:methodology} describes the
methodology used to answer the research questions; section~\ref{sec:results}
presents the results and section~\ref{sec:discussion} discusses their
implications; section~\ref{sec:threats}, threats to the validity of the study
are enumerated; section~\ref{sec:rw} describres the different work and existing
literature in the field of study; and, finally section~\ref{sec:conclusions}
concludes the main findings and elaborates on future work.
%
\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security expertise, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. It turns out that fixing these issues is as simple as
modifying the software codebase. However, these refactorings may have a negative
impact on software maintenance, mostly because developers follow the quickest
solution to fix the problem and not necessarily the most elegant and performant
one. Thus, we want to understand how security refactorings impact the
maintainability of software.

@@

~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
inserted to limit the number of bytes allowed in a DTLS handshake message for
\texttt{s}. This function replaces lines $18$---$21$.

~\ref{lst:func2} Checks if the maximum certificate list size
(\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
(\texttt{max\_len}). If that condition holds, the function returns the size of
the certificate list, otherwise it returns the maximum length.

~\ref{lst:func5} and~\ref{lst:func4} add the check to ensure that the message
length is never higher than the maxiumum allowed by the DTSL protocol in
different functions.

Although limiting the size of a handshake message seems an elementary problem to
repair, the example shows that there was a significant amount of changes
performed in the codebase which produced a negative impact in the project
maintainability. Even though the developer tries to simplify the codebase with
the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len} (~\ref{lst:func1}), it
also disrupts $2$ of the guidelines proposed by the Software Improvement Group
(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}. It adds a
new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
(~\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which contributes
to a potential violation of the
\emph{Keep Your Codebase Small} guideline. Hence, we have observed that the
impact of this refactoring on the OpenSSL maintainability was negative.

@@

There are security flaws that are more difficult to refactor than others. For
instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability with an \texttt{escape}
function. Knowing which patterns are more likely to increase maintainability
issues is one step forward to bring awareness to security engineers of what
patterns need more attention. The taxonomy of security patterns used to answer
this question is the same as the one provided by the authors of the dataset~\cite{Reis:2017:IJSSE}.
Maintainability is measured separately for each pattern.
%
\section{Methodology}\label{sec:methodology}
%
The methodology used to measure the impact of security refactorings on the
maintainability of open-source software is discussed this section. The methodology
comprehends the following steps, as illustrated in Figure~\ref{fig:met}:
\begin{enumerate}
	\item Extract relevant data from a dataset containing
$716$ security refactorings collected from open-source software available on
GitHub~\cite{Reis:2017:IJSSE}.
	\item A baseline of regular commits was
randomly collected from the list of projects of the main dataset to evaluate the impact of regular commits on the maintainability of open-source
software.
\item Use the Software Improvement Group's\footnote{SIG's website is available
at https://www.sig.eu/ (Accessed on \today{})} web-based source code analysis
service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
https://bettercodehub.com/ (Accessed on \today{})} to quantify
maintainability for both security and regular commits.
\end{enumerate}

%
\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
 	\caption{Study Methodology}
	\label{fig:met}
\end{figure}
%
\subsection{Dataset}
%
To evaluate the impact of security refactorings on the maintainability of
open-source software, we use a dataset of security flaws which is the outcome of
mining $94$ GitHub projects. Reis and Abreu (2017) mined open-source
software aiming at the extraction of real examples --- created by real
developers --- to test and assess the performance of static analysis tools since
using hand-seeded test cases or mutations could lead to misleading assessments
of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
dataset with $716$ test cases for $16$ security patterns. Some of the patterns 
are based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top 10 of 
2017~\cite{oswap:2017}. Each test case of the
dataset is a triplet: the commit before the refactoring, the commit responsible
for the refactoring, and the snippets of code that differ from one version to
another (typically, called \textit{diffs}) --- where one can easily review the
code used to fix the security flaw. In this study, we focus on computing the
maintainability of the commits before and after the security refactoring to
evaluate if the impact was positive, negative or none.

@@

@@

	\item \textbf{Denial-of-Service} Security flaws that can allow a cracker to
	flood or crashing services. This attacks usually occur when the system
	receives much traffic causing it to slow down and eventually stop. For
	example, flaws allowing crackers to trigger memory allocations corresponding
	to large length values (Listing~\ref{lst:vuln}).

	\item \textbf{Miscellaneous} This pattern comprises several others categories
	of security refactorings (e.g., \textit{path traversal and buffer overflow}).
	It contains several other security refactorings that do not have their own
	pattern yet. The patterns not satisfying the size requirement of more than 20
	refactorings are contemplated here: \textit{Resource Leak},
	\textit{SHA-1 Hash Function}, \textit{Broken Access Control}, \textit{Sensitive Data Exposure},
	\textit{Insufficient Attack Protection}, \textit{Overflow}, \textit{Path Traversal}.

@@

The dataset contemplates security flaws for more than 13 different languages
being PHP (39\%) and C (24\%) the most prevalent ones; see Figure~\ref{fig:lang}.
To classify the applications in the dataset, we use a taxonomy for open-source
software from previous work~\cite{7816479}. Figure~\ref{fig:domain}
presents the projects domain distribution: \textit{Aplication Software} ($25$),
software that provides end-users with functional systems; \textit{Web libraries
and frameworks} ($20$); \textit{System Software} ($18$), software that provides
services and infrustructures (e.g., operating systems, servers and databases);
\textit{Software Tool} ($13$), software that supports development (e.g.,
programming languages, compilers, package managers, IDEs); and, \textit{Non-web
libraries and frameworks} ($18$), for desktop and mobile software.
%
Table~\ref{tab:dataset} presents the descriptive statistics of the $94$
open-source projects involved in this study, including the number of forks, the number of
stars, the number of watchers, the number of contributors, the number of commits, the number of
branches, the number of releases,  size (cf. GitHub it is the size of the whole
repository, including all of its history, in kilobytes), the number of issues, and the
number of pull requests.

@@

As said before, the web-based source code analysis service \emph{Better Code
Hub} (BCH) is used to collect the maintanability reports of the refactorings of
each project. Table~\ref{tab:guidelines} presents the 10 guidelines
proposed by BCH's authors for delivering software that is easy to maintain~\cite{Visser:2016:OREILLY}. All guidelines are computed in the same evaluation by
BCH. During each guideline evaluation, the tool determines the compliance towards
one guideline by establishing limits for the percentage of code allowed to be in
each of the $4$ risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk}
and \emph{very high risk}). If the project does not violate the thresholds, then it
is compliant with the guideline. These thresholds are calibrated by BCH using
their own data/experience --- open-source and closed software systems. If a
project is compliant with a guideline, it means that it is at least $65\%$
better than the software used by BCH to calculate the thresholds\footnote{Check
the answer to \emph{How can I adjust the threshold for passing/not passing a
guideline?} at https://bettercodehub.com/docs/faq (Accessed on \today{})}.

Figure~\ref{fig:bchrep} shows an example of the report
provided by BCH for a project after finishing its evaluation. The examples
refers to the OpenSSL CVE-2014-3506 vulnerability refactoring ---
as described as motivating example in Section~\ref{sec:motivation}. This
version of OpenSSL only complies with 1 out of 10 guidelines: \emph{Write
Clean Code}.

@@

\emph{Units} are the smallest groups of code that can be maintained and executed
independently~\cite{Visser:2016:OREILLY}. McCabe~\cite{1702388} is a software
metric that calculates the number of linearly independent paths of the source
code. BCH uses this metric to measure the number of branch points in the
\emph{Write Simple Units of Code} guideline. The bar represents the top 30 of
units that violates the guideline, sorted by severity, which is indicated by the
colors of the checkboxes. The green bar represents the number of compliant
branch points per unit (\emph{at most 5}), i.e., the units are compliant with
ISO 25010~\cite{iso:2011}. Yellow, orange and red bars represent units that do
not comply with medium (\emph{above 5}), high (\emph{above 10}) and very high
(\emph{above 25}) severity levels. In the bar, there are marks that pinpoint the
compliance thresholds for each severity level. If the green mark is somewhere
in the green bar it means it is compliant with a low level of severity.

@@

Aiming to analyze the impact of security refactorings, we use BCH to compute
maintainability of two different versions of the project (Figure~\ref{fig:commit}):
\begin{itemize}
	\item $v_{s-1}$, the version containing the security flaw, i.e., before the
	refactoring;
	\item $v_{s}$, the version after fixing the security flaw, i.e., after the
	refactoring;
\end{itemize}

The BCH tool does not compute the final score that our study needs to compare
maintainability amongst different project versions. Thus, our work proposes an
equation to capture the distance between the current state of the project and
the standard thresholds based on previous work\cite{Olivari:2018}.
\begin{itemize}
	\item \textbf{Size project changes do not affect the maintainability
	difference between two versions, $\Delta M (v_{s-1},v_{s})$.} In this work, we
	aim to evaluate the alike security patterns occurring in different projects.
	Thus, normalization based on its size cannot be used - we convert relative
	data to the respective number of lines of code.
	\item \textbf{Distance to lower severity levels are less penalized than to the
	thresholds in high severity levels.} Severity level weights based on the
	severity level to count lines of code that violate maintainability guidelines.
\end{itemize}

@@

\noindent
where $G$ is the group of selected maintainability guidelines from BCH (Table~\ref{tab:guidelines}) and $v$ is the version of the software under evaluation.
The maintenance $M$ based on the guideline $g$ for a given version of a project
is computed with the following equation:

@@

\noindent
where $T(l)$ is the threshold in percentage of the lines of code that are
accepted to be non-compliant with the guideline for the severity level $l$. This
is a standard value defined by BCH. In other words, the factor $w$ is used in
Equation~\ref{eq:3} to highlight the lines of code that are not complying with
the guideline. Then, we compute the difference of maintainability between the
security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in
Figure~\ref{fig:commit}.

\subsection{Statistical Validation}\label{sec:statsval}
%
To validate the maintainability differences in different groups of commits
(e.g., baseline and security commits), we use the Paired Wilcoxon signed-rank
test with the significance level $\alpha = 0.05$~\cite{10.2307/3001968} . In
other words, we test the null hypothesis that the maintainability difference
between pairs of versions $v_{v-1}$, $v_v$ (i.e., before and after a
security-commit) follows a symmetric distribution around $0$. However, when the
difference between groups is zero, the observations are discarded. Thus, we use
a more robust approach of the wilcoxon test, the Wilcoxon test by Pratt which
provides an alternative that incorporates the cases where the maintainability
difference is zero~\cite{10.2307/2282543}. To understand the effect-size, as
advocated by the Common-language effect sizes\cite{graw:1992}, we compute the
mean difference, the median of the difference, and the percentage of cases that
reduce maintainability.
%
\section{Results}\label{sec:results}

In total, this study evaluates $607$ security and baseline commits from 94
different open-source projects distributed among 5 different main types of
software, Figure~\ref{fig:domain}. The following section presents the results
obtained for each research question.

@@

The impact of security and baseline commits in the maintainability of
open-source software is presented in Figure~\ref{fig:secvsreg}. For each type
of change, three types of impact of refactorings in the software maintainability
are presented: \emph{negative}, maintainability decreases (\emph{red}); \emph{none},
maintainability stays the same (\emph{yellow}); and, \emph{positive}, maintainability
increases (\emph{green}). In front of each group, it is presented the mean ($\overline{x}$)
and median ({$\widetilde{x}$) of the maintainability difference and the p-value
resulting from the Wilcoxon signed-rank test.

@@

In this study, we also evaluated the impact of each programming language on the
maintainability of open-source software, Figure~\ref{fig:lang_main}. Although the
dataset contains security refactorings for more than $13$ languages, we only
present the results for \emph{Java}, \emph{C}, \emph{Ruby} and \emph{PHP} because
one of the requirements of the Wilcoxon signed-rank test is that the input sample
has to have more than $20$ measurements. The other languages were integrated in the
\emph{Others} set. The results of the Wilcoxon signed-rank test by pattern returned
statistical evidence which supports that the maintainability of open-source software
decreases for security refactorings in 3 different languages: \emph{C}
(p = $5.72$x$10^{-9}$), Ruby (p = 0.057) and PHP (p = $3.54$x$10^{-6}$). The results
also provide statistical significance for when security refactorings are performed
in \emph{Java} (p = $3.61$x$10^{-2}$). \emph{Python} and the other languages
that integrate the \emph{Others} set contain more cases where the maintainability
increases. However, the results obtained for these patterns are not statistically
significant.

@@

The impact of each type of security refactoring on the maintainability of
open-source software is presented in Figure~\ref{fig:pat}. The results of the
Wilcoxon signed-rank test by pattern returned statistical evidence which
supports that the maintainability of open-source software decreases for 4
different patterns: Broken Autentication (p = 0.018), Denial-of-Service
(p = 0.029), Memory Leak (p = 0.009) and Miscellaneous (p = $3.07$x$10^{-4}$).
The results also returned statistical significance that evidence that maintainability
is more likely to not suffer differences for open-source software after removing
Components with Known Vulnerabilities (p = $7.96$x$10^{-4}$) and fixing Cross-Site
Scripting Vulnerabilities (p = 0.001). Injection and Cross-Site Request Forgery
produced more cases where the refactorings had a negative impact in the maintainability.
However, the results obtained for these patterns are not statistically significant.

@@

40.2\% of the security refactoring have a negative impact on the maintainability
of open-source software, Figure~\ref{fig:secvsreg}. This raises a new tradeoff when
developers need to fix vulnerabilities on their projects because they might be affecting
the software maintainability or even introducing new vulnerabilities. This study exhibits
proof that developers may have to reduce maintainability for the sake of safety. We understand
that developers should be able to produce secure code without affecting the maintainability of
their projects. But there are still some concerns that need to be addressed:
\begin{itemize}
	\item \textbf{Security Frameworks} need to be secure and user-friendly~\cite{5287006, 7676144}. Unfortunately,
	there are several vulnerabilities that stay undisclosed. Thus, developers should not
	trust them blindly and check defaults.
	\item \textbf{Documentation} of software should provide developers with the best practices to implement safe and maintainable software. Unmaintainable code is difficult to test, analyze and re-use.
	\item \textbf{Proper Authorization Mechanisms} should be designed and implemented. This is
	one of the most complex security features to implement. Some mechanisms might assist authorization maintenance in long run, such as the combination of \texttt{HTTP} filters with \texttt{Spring Security} annotations.
	\item \textbf{Deprecated code} should be avoided by developers. They should integrate tools such as BCH on their projects CI/CD pipelines to help them be more aware of the maintainability of their code and issues (e.g., dead and duplicated code).
	\item\textbf{Programming Languages} should provide new mechanisms to easily implement security patterns without endangering maintainability.
